---
title: "How to build, run and interrogate `cmdStan` models using the `stanUtils` package"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{stanUtils}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(blank = TRUE, width.cutoff = 75), message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>")
```

First ensure that `cmdStan` is installed on your computer.

```{r, results='hide'}
library(stanUtils)
```

# Set up

```{r}
# write example models
stan_codeA <- "data {int N; vector[N] y;} parameters {real mu;} model {y ~ normal(mu,1);} generated quantities {real mse; mse = mean((y - mu) .* (y - mu));}"
stan_codeB <- "data {int N; vector[N] y;} parameters {real mu; real<lower=0> sigma;} model {y ~ normal(mu,sigma);} generated quantities {real mse; mse = mean((y - mu) .* (y - mu));}"

cat(stan_codeA, file = "modelA.stan")
cat(stan_codeB, file = "modelB.stan")

# write *.ini file
inis <- list(mu = 10)
rstan::stan_rdump(list = ls(inis), file = "modelA.ini",  envir = list2env(inis))

inis <- list(mu = 10, sigma = 1)
rstan::stan_rdump(list = ls(inis), file = "modelB.ini",  envir = list2env(inis))

# write *.dat file
dat <- list(N = 100, y = rnorm(100, 10, 1))
rstan::stan_rdump(list = ls(dat), file = "modelA.dat",  envir = list2env(dat))
rstan::stan_rdump(list = ls(dat), file = "modelB.dat",  envir = list2env(dat))

# create simple makefile
cat("SRC_FILE=$(NAME).stan\n", file = "Makefile")
cat("MODEL_HOME=H:/CODE/packages/stanUtils/vignettes\n", file = "Makefile", append = TRUE)
cat("CMDSTAN_HOME=C:/cmdstan\n", file = "Makefile", append = TRUE)
cat("\n", file = "Makefile", append = TRUE)
cat("build: $(SRC_FILE)\n", file = "Makefile", append = TRUE)
cat("	cd $(shell cygpath -u ${CMDSTAN_HOME}) && make $(MODEL_HOME)/$(NAME).exe\n", file = "Makefile", append = TRUE)
cat("	rm -f $(NAME).hpp\n", file = "Makefile", append = TRUE)
cat("\n", file = "Makefile", append = TRUE)
cat("mcmc:\n", file = "Makefile", append = TRUE)
cat("	./$(NAME).exe sample algorithm=hmc num_samples=1000 num_warmup=1000 thin=1 init=$(NAME).ini data file=$(NAME).dat output file=$(NAME)$(chain).mcmc\n", file = "Makefile", append = TRUE)
cat("	$(RM) .RData\n", file = "Makefile", append = TRUE)

# build the model from the command line
system("make build NAME=modelA")
system("make build NAME=modelB")

# run the model
system("make mcmc NAME=modelA chain=1")
system("make mcmc NAME=modelA chain=2")
system("make mcmc NAME=modelB chain=1")
system("make mcmc NAME=modelB chain=2")
```

# Save model parameters and outputs

Parameters are kept as separate chains, whilst outputs are permutted.

```{r, echo = TRUE}
# extract and save mcmc outputs as *.rds file
(outA <- stan_extract(data = TRUE, mcmc = TRUE, model = "modelA", parameters = "mu", outputs = c("mu", "mse")))
stanUtils::stanSave(outA)

(outB <- stan_extract(data = TRUE, mcmc = TRUE, model = "modelB", parameters = c("mu", "sigma"), outputs = c("mu", "mse")))
stanUtils::stanSave(outB)

```

# Interrogate single model

Dianostic plots can be called directly to plot parameters. Alternatively the `posterior` function can be used to extract permutted chains for the outputs.

```{r}
# examine multiple chains in stanOutput object
traceplot(outA)
histplot(outA)

# examine permutted chains
(outpostA <- posterior(outA))
traceplot(outpostA)
histplot(outpostA)
```

# Interrogate multiple models

The `posterior` function can also be used to combined outputs from multple models.

```{r}
# examine permutted chains
(outpostAB <- posterior(models = c("modelA", "modelB"), pars = c("mu")))
traceplot(outpostAB)
histplot(outpostAB)
```


```{r, echo = FALSE, results = 'hide'}
files <- list.files(pattern = "[.]par");  file.remove(files)
files <- list.files(pattern = "[.]dat");  file.remove(files)
files <- list.files(pattern = "[.]ini");  file.remove(files)
files <- list.files(pattern = "[.]stan"); file.remove(files)
files <- list.files(pattern = "[.]exe");  file.remove(files)
files <- list.files(pattern = "[.]rds");  file.remove(files)
files <- list.files(pattern = "[.]mcmc"); file.remove(files)
file.remove("Makefile")
```
