---
title: "How to build, run and interrogate `cmdStan` models using the `stanUtils` package"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{stanUtils}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(blank = TRUE, width.cutoff = 75), message = FALSE, warning = FALSE, collapse = TRUE, comment = "# >")
```

This package is designed to make it easier to run and interrogate models using `cmdStan`, and the steps involved are detailed here. First ensure that `cmdStan` is installed on your computer, and then load the `stanUtils` package.

```{r, results='hide'}
library(stanUtils)
```

## Set up

Write the stan model code, data file, initial value file.

```{r}
# write example models
stan_codeA <- "data {int N; vector[N] y;} parameters {real mu;} model {y ~ normal(mu,1);} generated quantities {real mse; mse = mean((y - mu) .* (y - mu));}"
stan_codeB <- "data {int N; vector[N] y;} parameters {real mu; real<lower=0> sigma;} model {y ~ normal(mu,sigma);} generated quantities {real mse; mse = mean((y - mu) .* (y - mu));}"

cat(stan_codeA, file = "modelA.stan")
cat(stan_codeB, file = "modelB.stan")

# write *.ini file
inis <- list(mu = 10)
rstan::stan_rdump(list = ls(inis), file = "modelA.ini",  envir = list2env(inis))

inis <- list(mu = 10, sigma = 1)
rstan::stan_rdump(list = ls(inis), file = "modelB.ini",  envir = list2env(inis))

# write *.dat file
dat <- list(N = 100, y = rnorm(100, 10, 1))
rstan::stan_rdump(list = ls(dat), file = "modelA.dat",  envir = list2env(dat))
rstan::stan_rdump(list = ls(dat), file = "modelB.dat",  envir = list2env(dat))
```

```{r echo=FALSE}
cat("SRC_FILE=$(NAME).stan\n", file = "Makefile")
cat("MODEL_HOME=C:/Rpackages/stanUtils/vignettes\n", file = "Makefile", append = TRUE)
cat("CMDSTAN_HOME=C:/cmdstan\n", file = "Makefile", append = TRUE)
cat("\n", file = "Makefile", append = TRUE)
cat("build: $(SRC_FILE)\n", file = "Makefile", append = TRUE)
cat("	cd $(shell cygpath -u ${CMDSTAN_HOME}) && make $(MODEL_HOME)/$(NAME).exe\n", file = "Makefile", append = TRUE)
cat("	rm -f $(NAME).hpp\n", file = "Makefile", append = TRUE)
cat("\n", file = "Makefile", append = TRUE)
cat("mcmc:\n", file = "Makefile", append = TRUE)
cat("	./$(NAME).exe sample algorithm=hmc num_samples=1000 num_warmup=1000 thin=1 init=$(NAME).ini data file=$(NAME).dat output file=$(NAME)$(chain).mcmc\n", file = "Makefile", append = TRUE)
cat("	$(RM) .RData\n", file = "Makefile", append = TRUE)
```

We can also write a make file, for example:
```{r, echo = FALSE, comment=""}
# create simple makefile; e.g.
cat(readLines(con = "Makefile"), sep = "\n")

```

Build and run the model or models from the command line.

```{r}
# build the model from the command line
system("make build NAME=modelA")
system("make build NAME=modelB")

# run the model
system("make mcmc NAME=modelA chain=1")
system("make mcmc NAME=modelA chain=2")
system("make mcmc NAME=modelB chain=1")
system("make mcmc NAME=modelB chain=2")
```

## Save model parameters and outputs

For each model, extract the MCMC outputs into a `stanOuput` object using `stan_extract`, specifiying whether to treat the outputs as `parameters` (for which chain identities are retained) or `outputs` (for which the chains are permutted and combined). The specification of `parameters` and `outputs` can overlap, depending on downstream analyses. This flexibility is designed to keep the `stanOutput` object small, particularly for large model applications.

A call to `stanSave` will save the `stanOutput` object using `base::saveRDS`, with a file name that matches the model name. 

```{r, echo = TRUE}
# extract and save mcmc outputs as *.rds file
(outA <- stan_extract(model = "modelA", parameters = "mu", outputs = c("mu", "mse")))
stanSave(outA)

(outB <- stan_extract(model = "modelB", parameters = c("mu", "sigma"), outputs = c("mu", "sigma")))
stanSave(outB)

```
The `stanOuput` class object contains a variety of slots that are populated during a call to `stan_extract`:
```{r}
slotNames(outA)
```
In the current example, because it is of primary interest to users of `cmdStan`, we are concerned only with Bayesian outputs stored in `<object>@mcmc`.

## Interrogate single model

For each model, produce diagnostic plots through calls to `traceplot` and `histplot` which will plot `parameters`, keeping chains separate.

```{r}
# examine multiple chains in stanOutput object
traceplot(outA)
histplot(outA)
```

Alternatively, extract a posterior sample through a call to `posterior` which will extract the permutted chains for the specificed model `outputs` (provided they are stored in the `stanOutput` object). By default all `outputs` are extracted (i.e. those listed in `<object>@outputs`). These can be plotted, as well as printed.
```{r}
# examine permutted chains
(outpostA <- posterior(outA))
traceplot(outpostA)
histplot(outpostA)
print(outpostA)
```

## Interrogate multiple models

The `posteriors` function, when provided with one or more `stanPosterior` objects, can be used to combine `outputs` from multiple models into a single `stanPosteriors` object, which is useful for model comparison. Only outputs common to all models are retained.

```{r}
# examine permutted chains
(outpostAB <- posteriors(outpostA, posterior(outB)))
traceplot(outpostAB)
histplot(outpostAB)
print(outpostAB)
```

In addition, which may be important for particularly large models, a character vector of model names can also be used to combined outputs from multiple models into a single `stanPosteriors` object. This is achieved by directly reading the `*.rds` files created by `stanSave()`, possibly stored in different directories. In this case the `pars` argument must be specified. These can then also be plotted or printed as before.

```{r}
(outpostAB <- posteriors(models = c("modelA", "modelB"), pars = c("mu")))
```


```{r, echo = FALSE, results = 'hide'}
files <- list.files(pattern = "[.]par");  file.remove(files)
files <- list.files(pattern = "[.]dat");  file.remove(files)
files <- list.files(pattern = "[.]ini");  file.remove(files)
files <- list.files(pattern = "[.]stan"); file.remove(files)
files <- list.files(pattern = "[.]exe");  file.remove(files)
files <- list.files(pattern = "[.]rds");  file.remove(files)
files <- list.files(pattern = "[.]mcmc"); file.remove(files)
file.remove("USER_HEADER.hpp")
file.remove("Makefile")
```
